Multifunctional Lipid Probes Background

### 2025-01-30 IndividualStudies/AT_2025.qmd
updating with includes structure

## Study details

#### Authors
Alix Thomas, Rainer Mueller, Ana Kojic, Scotland E. Farley, Frank Stein, Per Haberkant, Carsten Schultz

#### Journal
Chemical Communications

<https://doi.org/10.1039/D4CC03599B>{target="_blank"}

#### Abstract
Phosphatidylethanolamine (PE) is a ubiquitous lipid species in higher eukaryotes which resides preferentially in mitochondria. Here, we synthesized a multifunctionalized PE derivative (1) de-signed to identify PE-binding proteins in intact cells through photo-crosslinking and subsequent isolation and proteomic analysis of the PE-protein conjugates. Due to its aromatic caging group, 1 is initially concentrating in perinuclear membranes. After uncaging, rapid translocation to mitochondria was observed. Hence, the tool is useful for tracking PE location and for determining the PE interactome. A trifunctional phosphatidic acid (PA) derivative in compari-son was rapidly metabolized and is hence more limited in its use.

#### Lipid probes utilized

#### Cell line analyzed

[HeLa](https://www.atcc.org/products/ccl-2){target="_blank"}

#### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

#### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

>>>>

{{< include /StudyInformation_includes/AT_2025_include.qmd >}}

:::

### 2025-01-30 IndividualStudies/DH_2017.qmd
updating with includes structure

## Study details

#### Authors

Doris Höglinger, André Nadler,  Per Haberkant, Joanna Kirkpatrick, Martina Schifferer, Frank Stein, Sebastian Hauke, Forbes D. Porter, Carsten Schultz

#### Journal

PNAS

<https://doi.org/10.1073/pnas.1611096114>{target="_blank"}

#### Abstract

Lipid-mediated signaling events regulate many cellular processes. Investigations of the complex underlying mechanisms are difficult because several different methods need to be used under varying conditions. Here we introduce multifunctional lipid derivatives to study lipid metabolism, lipid−protein interactions, and intracellular lipid localization with a single tool per target lipid. The probes are equipped with two photoreactive groups to allow photoliberation (uncaging) and photo–cross-linking in a sequential manner, as well as a click-handle for subsequent functionalization. We demonstrate the versatility of the design for the signaling lipids sphingosine and diacylglycerol; uncaging of the probe for these two species triggered calcium signaling and intracellular protein translocation events, respectively. We performed proteomic screens to map the lipid-interacting proteome for both lipids. Finally, we visualized a sphingosine transport deficiency in patient-derived Niemann−Pick disease type C fibroblasts by fluorescence as well as correlative light and electron microscopy, pointing toward the diagnostic potential of such tools. We envision that this type of probe will become important for analyzing and ultimately understanding lipid signaling events in a comprehensive manner.

#### Lipid probes utilized

Sphinganine

Diacylglycerol

8-3 Fatty acid

#### Cell line analyzed

[HeLa](https://www.atcc.org/products/ccl-2){target="_blank"}

#### Mass spectrometry quantification method

Peptide spectral counting

>>>>

{{< include /StudyInformation_includes/DH_2017_include.qmd >}}

### 2025-01-30 IndividualStudies/RM_2021.qmd
updating with includes structure

## Study details

#### Authors
Rainer Muller, Ana Kojic, Mevlut Citir, Carsten Schultz

#### Journal
Angewandte Chemie

<https://doi.org/10.1002/anie.202103599>{target="_blank"}

#### Abstract

We synthesized the first multifunctionalized phosphoinositide polyphosphate derivatives featuring a photo-removable protecting group (“cage”), a photo-crosslinkable diazirine group, and a terminal alkyne group useful for click chemistry. We demonstrate that the lipid derivatives readily enter cells. After photo-crosslinking, cell fixation and fluorescent tagging via click chemistry, we determined the intracellular location of the lipid derivatives before and after uncaging of the lipids. We find that there is rapid trafficking of PI(3,4)P2 and PI(3,4,5)P3 derivatives to the plasma membrane, opening the intriguing possibility that there is active transport of these lipids involved. We employed the photo-crosslinking and click chemistry functions to analyze the proteome of PI(3,4,5)P3-binding proteins. From the latter, we validated by RNAi that the putative lipid binding proteins ATP11A and MPP6 are involved in the transport of PI(3,4,5)P3 to the plasma membrane.

#### Lipid probes utilized

[PI(3,4,5)P~3~](/LipidProbe/PhosphatidylinositolTrisphosphate.qmd)

#### Cell line analyzed

[HeLa](https://www.atcc.org/products/ccl-2){target="_blank"}

#### Mass spectrometry quantification method

Peptide spectral counting

#### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

>>>>

{{< include /StudyInformation_includes/RM_2021_include.qmd >}}

:::

### 2025-01-30 IndividualStudies/SF_2024.qmd
updating with includes structure

## Study details

#### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

#### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

#### Abstract

Functions of the sphingolipids sphingosine and sphinganine in cells are not well established. While some signaling roles for sphingosine have been elucidated, the closely related sphinganine has been described only insofar as it does not elicit many of the same signaling responses. The underlying mechanisms behind the cell biological differences between these lipids are not well understood. Here, we prepared multifunctionalized derivatives of the two lipid species that only differ in a single double bond of the carbon backbone. Using these novel probes, we were able to define their spatiotemporal distribution within cells. Furthermore, we used these tools to systematically map the protein interactomes of both lipids. The lipid-protein conjugates, prepared through photo-crosslinking in live cells and extraction via click chemistry to azide beads, revealed significant differences in the captured proteins, highlighting their distinct roles in various cellular processes. This work elucidates mechanistic differences between these critical lipids and sets the foundation for further studies on the functions of sphingosine and sphinganine.

#### Lipid probes utilized

#### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

#### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

#### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

>>>>

{{< include /StudyInformation_includes/SF_2024_A_include.qmd >}}

### 2025-01-30 LipidProbe/1-10_FattyAcid.qmd
fleshing out overview
updating with includes structure

---
title: "Trifunctional 1-10 Fatty Acid"
---

@@

@@

#### Study details

##### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

##### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

##### Abstract

Functions of the sphingolipids sphingosine and sphinganine in cells are not well established. While some signaling roles for sphingosine have been elucidated, the closely related sphinganine has been described only insofar as it does not elicit many of the same signaling responses. The underlying mechanisms behind the cell biological differences between these lipids are not well understood. Here, we prepared multifunctionalized derivatives of the two lipid species that only differ in a single double bond of the carbon backbone. Using these novel probes, we were able to define their spatiotemporal distribution within cells. Furthermore, we used these tools to systematically map the protein interactomes of both lipids. The lipid-protein conjugates, prepared through photo-crosslinking in live cells and extraction via click chemistry to azide beads, revealed significant differences in the captured proteins, highlighting their distinct roles in various cellular processes. This work elucidates mechanistic differences between these critical lipids and sets the foundation for further studies on the functions of sphingosine and sphinganine.

##### Lipid probes utilized

1-10 Fatty acid

8-3 Fatty acid

##### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

##### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

##### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

>>>>

---
title: "Trifunctional 1-10 Fatty Acid"
crossref: 
  ref-hyperlink: true # (default is true)
---

@@

*Greater context to the distinction between the 8-3 Fatty Acid and 1-10 Fatty Acid probes in the [Diazirine Overview](/TrifunctionalLipidProbesOverview.qmd#sec-DiazirinePlacement)*

@@

{{< include /StudyInformation_includes/SF_2024_B_include.qmd >}}

:::

### 2025-01-30 LipidProbe/8-3_FattyAcid.qmd
updating with includes structure

#### Study details

##### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

##### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

##### Abstract

Functions of the sphingolipids sphingosine and sphinganine in cells are not well established. While some signaling roles for sphingosine have been elucidated, the closely related sphinganine has been described only insofar as it does not elicit many of the same signaling responses. The underlying mechanisms behind the cell biological differences between these lipids are not well understood. Here, we prepared multifunctionalized derivatives of the two lipid species that only differ in a single double bond of the carbon backbone. Using these novel probes, we were able to define their spatiotemporal distribution within cells. Furthermore, we used these tools to systematically map the protein interactomes of both lipids. The lipid-protein conjugates, prepared through photo-crosslinking in live cells and extraction via click chemistry to azide beads, revealed significant differences in the captured proteins, highlighting their distinct roles in various cellular processes. This work elucidates mechanistic differences between these critical lipids and sets the foundation for further studies on the functions of sphingosine and sphinganine.

##### Lipid probes utilized

1-10 Fatty acid

8-3 Fatty acid

##### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

##### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

##### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

@@

#### Study details

##### Authors

Doris Höglinger, André Nadler,  Per Haberkant, Joanna Kirkpatrick, Martina Schifferer, Frank Stein, Sebastian Hauke, Forbes D. Porter, Carsten Schultz

##### Journal

PNAS

<https://doi.org/10.1073/pnas.1611096114>{target="_blank"}

##### Abstract

Lipid-mediated signaling events regulate many cellular processes. Investigations of the complex underlying mechanisms are difficult because several different methods need to be used under varying conditions. Here we introduce multifunctional lipid derivatives to study lipid metabolism, lipid−protein interactions, and intracellular lipid localization with a single tool per target lipid. The probes are equipped with two photoreactive groups to allow photoliberation (uncaging) and photo–cross-linking in a sequential manner, as well as a click-handle for subsequent functionalization. We demonstrate the versatility of the design for the signaling lipids sphingosine and diacylglycerol; uncaging of the probe for these two species triggered calcium signaling and intracellular protein translocation events, respectively. We performed proteomic screens to map the lipid-interacting proteome for both lipids. Finally, we visualized a sphingosine transport deficiency in patient-derived Niemann−Pick disease type C fibroblasts by fluorescence as well as correlative light and electron microscopy, pointing toward the diagnostic potential of such tools. We envision that this type of probe will become important for analyzing and ultimately understanding lipid signaling events in a comprehensive manner.

##### Lipid probes utilized

Sphinganine

Diacylglycerol

8-3 Fatty acid

##### Cell line analyzed

[HeLa](https://www.atcc.org/products/ccl-2){target="_blank"}

##### Mass spectrometry quantification method

Peptide spectral counting

>>>>

{{< include /StudyInformation_includes/SF_2024_A_include.qmd >}}

:::

@@

{{< include /StudyInformation_includes/DH_2017_include.qmd >}}

### 2025-01-30 LipidProbe/PhosphatidicAcid.qmd
updating with includes structure

#### Study details

##### Authors
Alix Thomas, Rainer Mueller, Ana Kojic, Scotland E. Farley, Frank Stein, Per Haberkant, Carsten Schultz

##### Journal
Chemical Communications

<https://doi.org/10.1039/D4CC03599B>{target="_blank"}

##### Abstract
Phosphatidylethanolamine (PE) is a ubiquitous lipid species in higher eukaryotes which resides preferentially in mitochondria. Here, we synthesized a multifunctionalized PE derivative (1) de-signed to identify PE-binding proteins in intact cells through photo-crosslinking and subsequent isolation and proteomic analysis of the PE-protein conjugates. Due to its aromatic caging group, 1 is initially concentrating in perinuclear membranes. After uncaging, rapid translocation to mitochondria was observed. Hence, the tool is useful for tracking PE location and for determining the PE interactome. A trifunctional phosphatidic acid (PA) derivative in compari-son was rapidly metabolized and is hence more limited in its use.

##### Lipid probes utilized

##### Cell line analyzed

[HeLa](https://www.atcc.org/products/ccl-2){target="_blank"}

##### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

##### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

>>>>

{{< include /StudyInformation_includes/AT_2025_include.qmd >}}

### 2025-01-30 LipidProbe/Sphinganine.qmd
updating with includes structure

#### Study details

##### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

##### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

##### Abstract

Functions of the sphingolipids sphingosine and sphinganine in cells are not well established. While some signaling roles for sphingosine have been elucidated, the closely related sphinganine has been described only insofar as it does not elicit many of the same signaling responses. The underlying mechanisms behind the cell biological differences between these lipids are not well understood. Here, we prepared multifunctionalized derivatives of the two lipid species that only differ in a single double bond of the carbon backbone. Using these novel probes, we were able to define their spatiotemporal distribution within cells. Furthermore, we used these tools to systematically map the protein interactomes of both lipids. The lipid-protein conjugates, prepared through photo-crosslinking in live cells and extraction via click chemistry to azide beads, revealed significant differences in the captured proteins, highlighting their distinct roles in various cellular processes. This work elucidates mechanistic differences between these critical lipids and sets the foundation for further studies on the functions of sphingosine and sphinganine.

##### Lipid probes utilized

##### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

##### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

##### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

>>>>

{{< include /StudyInformation_includes/SF_2024_A_include.qmd >}}

### 2025-01-30 StudyInformation_includes/SF_2024_A_include.qmd
updating with includes structure

---
title: "Farley et al. 2024"
format: 
  html:
    embed-resources: true
standalone: true
---

```{r setup, echo=FALSE, include=FALSE}
## intitialize libraries
library(dplyr)
library(tidyr)
library(plotly)
library(ggplot2)
library(grid)
library(readr)
library(rmarkdown)
library(emojifont)
library(gt)
library(crosstalk)
library(DT)

## Opening the source file for the ggplot custom theme
source("ggplot_styles.R")

## Some data wrangling to clean up Scotty's data
SF_2024_data_raw <- readr::read_csv("DataTables/SF_ACS-ChemBio_2024_Limma_results_V3.csv")

SF_2024_data <- SF_2024_data_raw |>
	select(gene_name, keep_Huh7, comparison, logFC, pvalue, hit_annotation) |>
	filter(keep_Huh7 == TRUE) |>
	select(-keep_Huh7) |>
	mutate(LipidProbe = case_when(comparison == "FA_8_3_UV - FA_8_3_noUV" ~ "8-3 FA",
															 comparison == "FA_1_10_UV - FA_1_10_noUV" ~ "1-10 FA",
															 comparison == "Spa_UV - Spa_noUV" ~ "Spa",
															 comparison == "Sph_UV - Sph_noUV" ~ "Sph")) |>
	filter(LipidProbe == "8-3 FA" | LipidProbe == "1-10 FA" | LipidProbe == "Spa" | LipidProbe == "Sph") |>
	select(-comparison) |>
	glimpse()

# Organize the hit annotations as factors
SF_2024_data$hit_annotation <- factor(pull(SF_2024_data, hit_annotation), levels = c("enriched hit", "hit", "enriched candidate", "candidate", "no hit", "NA"), ordered = TRUE)

# Organize the probes as factors
SF_2024_data$LipidProbe <- factor(pull(SF_2024_data, LipidProbe), levels = c("Spa", "Sph", "8-3 FA", "1-10 FA"), ordered = TRUE)

# Building an ID column to maintain the order on the x-axis
ID <- c()
for(i in 0:(nrow(SF_2024_data)/4)){
	ID[i] = i
}

# Applies the ID column to the 1-10 FA results
SF_2024_ranked_1_10_FA <- SF_2024_data |>
    arrange(logFC) |>
	filter(LipidProbe == "1-10 FA") |>
	mutate(ID = ID)

# Applies the ID column to the 8-3 FA results
SF_2024_ranked_8_3_FA <- SF_2024_data |>
    arrange(logFC) |>
	filter(LipidProbe == "8-3 FA") |>
	mutate(ID = ID)

# Applies the ID column to the Spa results
SF_2024_ranked_Spa <- SF_2024_data |>
    arrange(logFC) |>
	filter(LipidProbe == "Spa") |>
	mutate(ID = ID)

# Applies the ID column to the Sph results
SF_2024_ranked_Sph <- SF_2024_data |>
    arrange(logFC) |>
	filter(LipidProbe == "Sph") |>
	mutate(ID = ID)

# Assembles the dataset to plot
SF_2024_ranked <- rbind(SF_2024_ranked_Sph, SF_2024_ranked_Spa, SF_2024_ranked_8_3_FA, SF_2024_ranked_1_10_FA)

unique(SF_2024_data$hit_annotation)

print(head(SF_2024_data))
```

## Publication information

**Title:** Trifunctional Sphinganine: A New Tool to Dissect Sphingolipid Function

**Authors:** Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz*

(* corresponding: <schulcar@ohsu.edu>{target="_blank"})

**Date published:**

**Journal, ISSN:** [ACS Chemical Biology](https://pubs.acs.org/doi/10.1021/acschembio.3c00554){target="_blank"}, 1554-8929, 1554-8937

**DOI:** <https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

## Abstract

@@

## Data visualization

#### Ranked-order Log2 Fold-change plots

:::{.border}
```{r, echo=FALSE}
#| fig-cap: ""

RankedOrderPlotStandard(SF_2024_ranked)
```
:::

#### Volcano plots

:::{.border}
```{r, echo=FALSE}
#| fig-cap: ""

VolcanoPlotStandardized(SF_2024_data)

```
:::

## Enriched Proteins

:::{.border}
```{r, echo=FALSE, warning=FALSE}
### Here is the code to make the filterable lists shown in the crosstalk table below

# Creates a reactive dataframe
SF_2024_data_reactive <- crosstalk::SharedData$new(SF_2024_data)

filter_select(id = "LipidProbe",
				label = "Functional probe used in pulldown:", 
				sharedData = SF_2024_data_reactive,
				group = ~LipidProbe)

filter_select(id = "hit_annotation",
				label = "Enrichment criteria:", 
				sharedData = SF_2024_data_reactive,
				group = ~hit_annotation)

datatable(SF_2024_data_reactive)
```
:::

>>>>

## Study details

#### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

#### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

#### Abstract

@@

#### Lipid probes utilized

#### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

#### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

#### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
